---
title: "sc_lung_3 lung carcinomas"
output:
  pdf_document: default
  html_document: default
date: "2023-05-23"
---

#BIG NOTE:
#ALL THE GENES ARE IN ENSEMBL NOMENCLATURE, EXCEPT THE MITOCONDRIAL GENES
#THOSE ARE IN THE GENE SYMBOLS NOMENCLATURE

#libraries
```{r}
library(Seurat)
library(SeuratData)
library(gridExtra)
library(tidyverse)
library(data.table)
library(SingleCellExperiment)
library(scds)
library(useful)
library(ggplot2)

install.packages("Seurat")
install.packages("org.Hs.eg.db")
library(Seurat)
library(org.Hs.eg.db)
```

#DETERMINE THE DIRECTORY:
```{r}
knitr::opts_knit$set(root.dir = "C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/")

```


#load the data:
https://www.ebi.ac.uk/gxa/sc/experiments/E-MTAB-6653/downloads 
#1º:Load the cells
```{r}
library(readr)
mtx_cols <- read.table("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/barcodes.mtx_cols", header = FALSE, sep = "\t")
View(mtx_cols)

barcodes<-mtx_cols$V1
barcodes<-as.data.frame(barcodes,header=FALSE,rownames = FALSE)
file_path_<- file.path("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/", "barcodes.tsv.gz")
write.table(barcodes,file_path_,sep="\t",quote = FALSE,row.names = FALSE,col.names = FALSE)
View(barcodes)
```

#2º:Load the genes, convert into tsv and compress into .gz
```{r}
mtx_rows <- read.table("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/features.mtx_rows", header = FALSE, sep = "\t")
mtx_rows<-as.data.frame(mtx_rows$V1)
View(mtx_rows)

genes<-mtx_rows$`mtx_rows$V1`
genes<-as.data.frame(genes,header=FALSE,rownames = FALSE)
file_path <- file.path("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/", "features.tsv.gz")
write.table(genes,file_path,sep="\t",quote = FALSE,row.names = FALSE,col.names = FALSE)
View(genes)
```

#3º:TRY TO CONVERT THE ENSEMBL GENES INTO GENE SYMBOL AND DETERMINE THE MITOCONDRIAL GENES:
```{r}
#installations missing:
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("biomaRt")

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("org.Hs.eg.db")
library(ggplot2)
library(Matrix)
library(Seurat)
library(patchwork)
library(dplyr)
library(biomaRt)
library(org.Hs.eg.db)

#input list of the genes in the genes.tsv:
data_genes<-genes$genes

#BiomaRt method:
listEnsembl()
ensembl_<-useEnsembl(biomart = "genes")
datasets<-listDatasets(ensembl_)
ensembl <- useMart("ensembl", dataset="hsapiens_gene_ensembl")
attr<-listAttributes(ensembl)#here we define the reference to which type of conversion type we want to convert the ensembl genes
bm <- getBM(attributes=c("ensembl_gene_id", "external_gene_name"), values=data_genes, mart=ensembl)

# Check mitochondrial genes before conversion
print("Mitochondrial genes before conversion:")
print(grep("^MT-", data_genes, value = TRUE))#NO MITOCONDRIAL GENES BEFORE CONVERSION

#DETERMINE THE GENE SYMBOLS:
gene.symbols <- bm$external_gene_name[match(data_genes, bm$ensembl_gene_id)]
features_<-as.matrix(cbind(genes$genes,gene.symbols))
colnames(features_)<-c("ensembl_genes","gene_symbols")

#CREATE A NEW DATAFRAME WITH ENSEMBL GENES AND THE MITOCONDRIAL GENES IN GENE.SYMBOLS:
features<-as.data.frame(features_)
MT_GENES<-print(grep("^MT-",features_, value = TRUE))#13 MITOCONDRIAL GENES AFTER CONVERSION
MT_GENES<-as.list(MT_GENES)
features<- features[match(features$gene_symbols, MT_GENES, nomatch = 0) != 0,]

#CREATE THE FINAL FEATURES DATAFRAME:
features_df<-as.data.frame(features_)
#DO ONE BY ONE:
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198695"]<-"MT-ND6"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198712"]<-"MT-CO2"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198727"]<-"MT-CYB"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198763"]<-"MT-ND2"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198786"]<-"MT-ND5"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198804"]<-"MT-CO1"#
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198840"]<-"MT-ND3"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198886"]<-"MT-ND4"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198888"]<-"MT-ND1"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198899"]<-"MT-ATP6"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198938"]<-"MT-CO3"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000212907"]<-"MT-ND4L"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000228253"]<-"MT-ATP8"


View(features_df)
#CREATE THE TSV FILE WITH GENES CALLED FEATURES.TSV
file_path <- file.path("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/", "features.tsv.gz")
write.table(features_df,file_path,sep="\t",quote = FALSE,row.names = FALSE,col.names = FALSE)
#1823 genes-not converted.gz
#22156 genes from the 23979 were actually converted

mtx_matrix<- read.table("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/matrix.mtx", header = FALSE, sep = "\t")
mtx_rows<-as.data.frame(mtx_rows$V1)
View(mtx_matrix)
file_path <- file.path("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/", "matrix.mtx.gz")
write.table(mtx_matrix,file_path,sep="\t",quote = FALSE,row.names = FALSE,col.names = FALSE)

```


#4º:ALL THE DATASET IN THE ACCORDING FORMATS ACCEPTED BY THE FUNCTION READ10X
```{r}
#gene.column = 1
lung.data <- Read10X(data.dir ="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/")
                     #,gene.column = 2,unique.features=TRUE)#try to change the genes here in the original data 23,979

feature_names <- rownames(lung.data)

# Identify features to keep (those without a dot followed by numbers)
features_to_keep <- !grepl("\\.\\d+", feature_names)

# Subset the Seurat object
lung_filtered <- lung.data[features_to_keep,]

#in the case, of gene names
lung_filtered <- lung.data[rownames(lung.data) != "",]

View(lung_filtered)
View(as.data.frame(lung_filtered))

#CREATE THE SEURAT OBJECT WITH THE GENE NAMES AND MINIMUM OF 3 CELLS AND 200 GENES:
lung<- CreateSeuratObject(counts = lung_filtered, project = "lung", min.cells = 3, min.features = 200)

dim(lung_data)
lung<-lung_data

#save para o diogo:
saveRDS(lung_data,file="C:/Users/ritas/Desktop/single_cell_Rita/lung_data.RDS")

View(lung)
head(rownames(lung),20)
View(lung@assays$RNA@counts@Dimnames)

############################################################################
#to join the cells automatic annotation to the data:
signature_matrix<-lung@assays[["RNA"]]@counts
write.table(signature_matrix, file="signature_matrix.txt", row.names=TRUE, col.names=TRUE,sep = "\t",quote = FALSE)
df = read.table("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/signature_matrix.txt")
#23898 genes with gene names
############################################################################

```


#DETERMINE THE PERCENTAGE OF MITOCONDRIAL GENES ACCORDING TO THE PATTERN:
```{r}
#in the case of mouse genes
lung[["percent.mt"]] <- PercentageFeatureSet(lung, pattern = "^mt-") 
#OR
#in the case of human genes
lung[["percent.mt"]] <- PercentageFeatureSet(lung, pattern = "^MT-") 
lung[["percent.rp"]] <- PercentageFeatureSet(lung, pattern = "^RP(L|S)") 


VlnPlot(lung,features = c("nFeature_RNA","nCount_RNA","percent.mt","percent.rp"),ncol=4)

```

#OVERVIEW OF THE NFEATURE_RNA AND NCOUNT_RNA AND PERCENT.MT:
```{r}
View(lung@meta.data)
FeatureScatter(lung,feature1 = "nCount_RNA",feature2 = "nFeature_RNA")+geom_smooth(method = "lm")

#COMPARING PLOTS:
plot1 <- FeatureScatter(lung, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(lung, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
plot2
```

#FILTERING:
```{r}
lung<-subset(lung,subset=nFeature_RNA>200 & nFeature_RNA<5000 & percent.mt<25 & nCount_RNA<40000 & percent.rp<65)
#FIND DUBLETS:    [SOME AREAS THAT WERE MORE SEQUENCED, IN DUPLICATES]
lung_sce <- as.SingleCellExperiment(lung)
lung_sce <- cxds_bcds_hybrid(lung_sce, estNdbl=TRUE)#Function to find doublets/multiples in UMI
lung<- AddMetaData(
  object = lung,
  metadata = lung_sce$hybrid_call,
  col.name = "doublet"
)

table(lung_sce$hybrid_call)
lung<- subset(lung, subset = doublet == FALSE)

View(lung_sce)

#FALSE:30148 
#TRUE:1163   DOUBLETS
```

#3.Normalize data
```{r}
#1º:SEE THE MEANS AND VARIANCE OF THE DATA
lung<-lung_data
means <- sparseMatrixStats::rowMeans2(lung@assays$RNA@layers$counts)
variance <- sparseMatrixStats::rowVars(lung@assays$RNA@layers$counts)
plot(means, variance)
```

#2-A.TRY TO NORMALIZE THE DATA WITH LOG NORMALIZE:
```{r}
lung.norm<- NormalizeData(lung, normalization.method = "LogNormalize",scale.factor = 10000)

means <- sparseMatrixStats::rowMeans2(as.matrix(lung.norm@assays$RNA@layers$data))
variance <- sparseMatrixStats::rowVars(lung.norm@assays$RNA@layers$data)
plot(means, variance)

```

#2-B.TRY TO NORMALIZE THE DATA WITH SCTransform:
#NOTE:THE SCTRANSFORM TOOK A LONG TIME AND LEAD THE PC DOWN, SO I OPTED TO GO WITH THE CLASSIC NORMALIZATION:
```{r}
lung_sctranform <- SCTransform(lung,verbose=FALSE)

means_sctranform<- sparseMatrixStats::rowMeans2(lung_sctranform@assays$SCT@data)
variance_sctranform<- sparseMatrixStats::rowVars(lung_sctranform@assays$SCT@data)
plot(means_sctranform, variance_sctranform)

```


#3.SCALING (optional)------------------------------
```{r}
hist(sparseMatrixStats::colSums2(lung.norm@assays$RNA@layers$data), main = "Counts per Cell")
```

#4.IDENTIFY HIGHLY VARIABLE FEATURES (GENES)---------------------------------------
```{r}
nsclc <- FindVariableFeatures(lung.norm)

top10 <- head(VariableFeatures(nsclc), 10)

plot1 <- VariableFeaturePlot(nsclc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
plot2
plot1

```
#Identify the 10 most highly variable genes:
```{r}
top10<-head(VariableFeatures(nsclc),10)
top100<-head(VariableFeatures(nsclc),100)
print(top10)
print(top100)
```

#Plot variable features with and without labels:
```{r}
plot1<-VariableFeaturePlot(nsclc)
LabelPoints(plot=plot1,points = top10,repel = TRUE)#REPEL=TRUE, SHOWS THET POINTS WELL DISTINGUISH BETWEEN EACH OTHER.
```
##5.SCALING
REMOVE THE DESVIATIONS THAT LEAD TO THE CLUSTER OF THE CELLS, CAUSED BY THE DIFFERENCES ON CELL CYCLE OF THIS CELLS (BIOLOGICAL PROBLEM) OR THE BATCH EFFECT (TECNIQUAL PROBLEM); AND NOT BY THE BIOLOGICAL EFFECT.

```{r}
all.genes<-rownames(nsclc)
nsclc.scaled<-ScaleData(nsclc,features = all.genes)
str(nsclc.scaled)
#notes:
#normalized counts is stored in the data slot
#scaling data stored in scale.data
```
#B.DOWNSTREAM ANALYSIS
#6.Perform Linear dimensionality reduction------------------------------------------------------------------------------
```{r}
#6-A:TO PCA dimensionality reduction
ElbowPlot(lung)
lung<-RunPCA(nsclc.scaled,features = VariableFeatures(object = nsclc.scaled))
print(lung[["pca"]],dims=1:5,nfeatures=5)

VizDimLoadings(lung, dims = 1:2, reduction = "pca")

DimPlot(lung, reduction = "pca") + NoLegend()

FeaturePlot(lung,reduction = "pca",dims = c(1,2),features="percent.rp")

DimHeatmap(lung,dims=1,cells=500,balanced = TRUE)
DimHeatmap(lung,dims=1:10,cells=500,balanced = TRUE,nfeatures=50)
DimHeatmap(lung,dims=1:15,cells=500,balanced = TRUE)

```
```{r}
#6-B:FindNeighbors
#PARAMETERS:
#k.param: defines k for the k-nearest neighbor algorithm (depends- not a right number of k)
#reduction: Reduction to use as input for building the (S)NN
ElbowPlot(lung)

lung <- FindNeighbors(lung, dims = 1:20)

```

```{r}
#6-C:FindClusters
#PARAMETERS:
#RESOLUTION: Value of the resolution parameter, use a value above (below) 1.0 if you want to obtain a larger (smaller) number of communities.
#ex: resolution=0.5- 24 communities; resolution=0.8- 7communities
lung <- FindClusters(lung,resolution = seq(0.2, 0.7, 0.1))
#21 clusters

dim(lung)

# Look at cluster IDs of the first 5 cells
head(Idents(lung), 5)#obtain 34 levels
```

```{r}
#6-D:Plot the clusters:
p <- ggplot(data=as.data.table(lung@meta.data)[,.N,by="seurat_clusters"], mapping = aes(x=seurat_clusters, y=N, fill=seurat_clusters)) +
  geom_bar(stat="identity") +
  labs(x="Cluster Number", y="Number of Cells in that Cluster") +
  theme_classic()
print(p)
```

#UMAP VISUALIZATION:
```{r}
lung <- RunUMAP(lung, dims = 1:20)


#PLOT DIM:
#pt.size: Adjust point size for plotting
#shuffle: Whether to randomly shuffle the order of points. This can be useful for crowded plots if points of interest are being buried. (default is FALSE)
DimPlot(lung, reduction = "umap", label = TRUE,group.by = c("RNA_snn_res.0.2","RNA_snn_res.0.3","RNA_snn_res.0.4","RNA_snn_res.0.5","RNA_snn_res.0.6","RNA_snn_res.0.7"), label.box = FALSE)  & NoLegend() & NoAxes()

DimPlot(lung, reduction = "umap")
DimPlot(lung, reduction = "umap", pt.size = NULL, shuffle = TRUE) + coord_fixed()

Idents(object = lung) <-lung@meta.data$RNA_snn_res.0.6

FeaturePlot(lung, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rp"), order = TRUE) & NoLegend() & NoAxes()
```

#new_step:
```{r}
FindClusters(object = lung)
#AUTOMATIC ANNOTATION WITH BlueprintEncodeData()
library(SingleR)

# predict cell types based on the reference
lung_counts<-GetAssayData(lung, slot="counts")

pred <- SingleR(test=lung_counts, ref=ref, labels=ref$label.main)

lung$singleR.labels<-pred$labels[match(rownames(lung@meta.data),rownames(pred))]
DimPlot(lung,reduction = "umap",group.by = "singleR.labels")

table(pred$labels)
lung$singleR_annot = pred$labels
lung@assays$RNA@counts@Dimnames[[2]]=lung$singleR_annot
View(lung)

#Annotation diagnosis:
pred
pred$scores

plotScoreHeatmap(pred)
plotDeltaDistribution(pred)
#get reference data
ref<-celldex::HumanPrimaryCellAtlasData()
x<-as.data.frame(colData(ref))

#run singleR(default mode)
lung_counts<-GetAssayData(lung, slot="counts")

BiocManager::install("SingleR")
library(SingleR)
pred<-SingleR(test=lung_counts,ref=ref,labels = ref$label.main)

lung$singleR.labels<-pred$labels[match(rownames(lung@meta.data),rownames(pred))]
DimPlot(lung,reduction = "umap",group.by = "singleR.labels")

#Annotation diagnosis:
pred
pred$scores

plotScoreHeatmap(pred)
plotDeltaDistribution(pred)

#Comparing unsupervised clustering
tab<-table(Assigned=pred$labels,Clusters=lung$seurat_clusters)
library(pheatmap)
pheatmap(log10(tab+10),color=colorRampPalette(c("white","blue"))(10))
```




#7-Finding differentially expressed features (cluster biomarkers)
```{r}
# find all markers of cluster 2
cluster2.markers <- FindMarkers(lung, ident.1 = 2)
head(cluster2.markers, n = 5)

cluster5.markers <- FindMarkers(lung, ident.1 = 5, ident.2 = c(0, 3))
cluster5.markers

# find markers for every cluster compared to all remaining cells, report only the positive
# ones
lung.markers <- FindAllMarkers(lung)

#SAVE THE SEURAT OBJECT:
saveRDS(lung, file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/lung.RDS")

lung<-readRDS(file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/lung.RDS")
View(lung)

library(Seurat)
lung_combined.markers <- FindAllMarkers(lung, only.pos = FALSE,min.pct=0.1)
View(lung_combined.markers)

write.table(lung_combined.markers,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/AllMarkerslung.txt",quote = FALSE,row.names = FALSE,col.names = TRUE,sep = "\t")

#READ THE FILE WITH ALL MARKERS DIFFERENTIAL EXPRESSED:
k<-read.table(file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/AllMarkerslung.txt",header = TRUE)
View(k)

library(dplyr)
#OBTAIN ONLY THE DIFFERENTIAL EXPRESSED GENES:
b<-k %>% 
  group_by(cluster) %>% dplyr::filter(p_val_adj<0.05)
View(b_0)  

#automatic function azimuth:
devtools::install_github("satijalab/seurat", "seurat5")
devtools::install_github("satijalab/seurat-data", "seurat5")
devtools::install_github("satijalab/azimuth", "seurat5")
library(seurat5)
library(Seurat)
library(Azimuth)
library(SeuratData)
library(patchwork)

lung_annotation <- RunAzimuth(lung, reference ="Lung v2 (HLCA)")

#OBTAIN THE LIST OF DIFFERESNTIAL EXPRESSED GENES PER CLUSTER:
#A.CLUSTER 0:
b_0<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="0")

lung_cluster0_combined.markers<-as.data.frame(b_0)
lung_cluster0_combined.markers$diff_pct<-(lung_cluster0_combined.markers$pct.1-lung_cluster0_combined.markers$pct.2)
PLOT_CLUSTER_0<-FeaturePlot(lung,features = c("MCEMP1","UPP1", "HLA-DQA1", "C5AR1", "HLA-DMA", "AIF1", "MRC1", "CCL18"))

genes_cluster0 <- b_0$gene
View(lung_cluster0_combined.markers)

write.table(genes_cluster0,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster0.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 1:
b_1<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="1")
View(b_1)

lung_cluster1_combined.markers<-as.data.frame(b_1)
genes_cluster1 <- b_1$gene
View(genes_cluster1)

write.table(genes_cluster1,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster1.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 2:
b_2<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="2")
View(b_2)

lung_cluster2_combined.markers<-as.data.frame(b_2)
genes_cluster2 <- b_2$gene
View(genes_cluster2)

write.table(genes_cluster2,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster2.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 3:
b_3<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="3")
View(b_3)

lung_cluster3_combined.markers<-as.data.frame(b_3)
genes_cluster3 <- b_3$gene
View(genes_cluster3)

write.table(genes_cluster3,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster3.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 4:
b_4<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="4")
View(b_4)

lung_cluster4_combined.markers<-as.data.frame(b_4)
genes_cluster4 <- b_4$gene
View(genes_cluster4)

write.table(genes_cluster4,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster4.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 5:
b_5<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="5")
View(b_5)

lung_cluster5_combined.markers<-as.data.frame(b_5)
genes_cluster5 <- b_5$gene
View(genes_cluster5)

write.table(genes_cluster5,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster5.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 6:
b_6<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="6")
View(b_6)

lung_cluster6_combined.markers<-as.data.frame(b_6)
genes_cluster6 <- b_6$gene
View(genes_cluster6)

write.table(genes_cluster6,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster6.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 7:
b_7<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="7")
View(b_7)

lung_cluster7_combined.markers<-as.data.frame(b_7)
genes_cluster7 <- b_7$gene
View(genes_cluster7)

write.table(genes_cluster7,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster7.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 8:
b_8<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="8")
View(b_8)

lung_cluster8_combined.markers<-as.data.frame(b_8)
genes_cluster8 <- b_8$gene
View(genes_cluster8)

write.table(genes_cluster8,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster8.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 9:
b_9<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="9")
View(b_9)

lung_cluster9_combined.markers<-as.data.frame(b_9)
genes_cluster9 <- b_9$gene
View(genes_cluster9)

write.table(genes_cluster9,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster9.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 10:
b_10<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="10")
View(b_10)

lung_cluster10_combined.markers<-as.data.frame(b_10)
genes_cluster10 <- b_10$gene
View(genes_cluster10)

write.table(genes_cluster10,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster10.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 11:
b_11<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="11")
View(b_11)

lung_cluster11_combined.markers<-as.data.frame(b_11)
genes_cluster11 <- b_11$gene
View(genes_cluster11)

write.table(genes_cluster11,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster11.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 12:
b_12<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="12")
View(b_12)

lung_cluster12_combined.markers<-as.data.frame(b_12)
genes_cluster12 <- b_12$gene
View(genes_cluster12)

write.table(genes_cluster12,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster12.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 13:
b_13<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="13")
View(b_13)

lung_cluster13_combined.markers<-as.data.frame(b_13)
genes_cluster13 <- b_13$gene
View(genes_cluster13)

write.table(genes_cluster13,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster13.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 14:
b_14<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="14")
View(b_14)

lung_cluster14_combined.markers<-as.data.frame(b_14)
genes_cluster14 <- b_14$gene
View(genes_cluster14)

write.table(genes_cluster14,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster14.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 15:
b_15<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="15")
View(b_15)

lung_cluster15_combined.markers<-as.data.frame(b_15)
genes_cluster15 <- b_15$gene
View(genes_cluster15)

write.table(genes_cluster15,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster15.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 16:
b_16<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="16")
View(b_16)

lung_cluster16_combined.markers<-as.data.frame(b_16)
genes_cluster16 <- b_16$gene
View(genes_cluster16)

write.table(genes_cluster16,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster16.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 17:
b_17<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="17")
View(b_17)

lung_cluster17_combined.markers<-as.data.frame(b_17)
genes_cluster17 <- b_17$gene
View(genes_cluster17)

write.table(genes_cluster17,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster17.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 18:
b_18<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="18")
View(b_18)

lung_cluster18_combined.markers<-as.data.frame(b_18)
genes_cluster18 <- b_18$gene
View(genes_cluster18)

write.table(genes_cluster18,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster18.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 19:
b_19<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="19")
View(b_19)

lung_cluster19_combined.markers<-as.data.frame(b_19)
genes_cluster19 <- b_19$gene
View(genes_cluster19)

write.table(genes_cluster19,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster19.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 20:
b_20<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="20")
View(b_20)

lung_cluster20_combined.markers<-as.data.frame(b_20)
genes_cluster20 <- b_20$gene
View(genes_cluster20)

write.table(genes_cluster20,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster20.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 21:
b_21<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="21")
View(b_21)

lung_cluster21_combined.markers<-as.data.frame(b_21)
genes_cluster21 <- b_21$gene
View(genes_cluster21)

write.table(genes_cluster21,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster21.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 22:
b_22<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="22")
View(b_22)

lung_cluster22_combined.markers<-as.data.frame(b_22)
genes_cluster22 <- b_22$gene
View(genes_cluster22)

write.table(genes_cluster22,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster22.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 23:
b_23<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="23")
View(b_23)

lung_cluster23_combined.markers<-as.data.frame(b_23)
genes_cluster23 <- b_23$gene
View(genes_cluster23)

write.table(genes_cluster23,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster23.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 24:
b_24<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="24")
View(b_24)

lung_cluster24_combined.markers<-as.data.frame(b_24)
genes_cluster24 <- b_24$gene
View(genes_cluster24)

write.table(genes_cluster24,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster24.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 25:
b_25<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="25")
View(b_25)

lung_cluster25_combined.markers<-as.data.frame(b_25)
genes_cluster25 <- b_25$gene
View(genes_cluster25)

write.table(genes_cluster25,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster25.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 26:
b_26<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="26")
View(b_26)

lung_cluster26_combined.markers<-as.data.frame(b_26)
genes_cluster26 <- b_26$gene
View(genes_cluster26)

write.table(genes_cluster26,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster26.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 27:
b_27<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="27")
View(b_27)

lung_cluster27_combined.markers<-as.data.frame(b_27)
genes_cluster27 <- b_27$gene
View(genes_cluster27)

write.table(genes_cluster27,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster27.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 28:
b_28<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="28")
View(b_28)

lung_cluster28_combined.markers<-as.data.frame(b_28)
genes_cluster28 <- b_28$gene
View(genes_cluster28)

write.table(genes_cluster28,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster28.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 29:
b_29<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="29")
View(b_29)

lung_cluster29_combined.markers<-as.data.frame(b_29)
genes_cluster29 <- b_29$gene
View(genes_cluster29)

write.table(genes_cluster29,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster29.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 30:
b_30<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="30")
View(b_30)

lung_cluster30_combined.markers<-as.data.frame(b_30)
genes_cluster30 <- b_30$gene
View(genes_cluster30)

write.table(genes_cluster30,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster30.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 31:
b_31<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="31")
View(b_31)

lung_cluster31_combined.markers<-as.data.frame(b_31)
genes_cluster31 <- b_31$gene
View(genes_cluster31)

write.table(genes_cluster31,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster31.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 32:
b_32<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="32")
View(b_32)

lung_cluster32_combined.markers<-as.data.frame(b_32)
genes_cluster32 <- b_32$gene
View(genes_cluster32)

write.table(genes_cluster32,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster32.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#CLUSTER 33:
b_33<-b %>% 
  group_by(cluster) %>% dplyr::filter(cluster=="33")
View(b_33)

lung_cluster33_combined.markers<-as.data.frame(b_33)
genes_cluster33 <- b_33$gene
View(genes_cluster33)

write.table(genes_cluster33,file="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/MARKERS_PER_CLUSTER/genes_cluster33.txt",row.names = FALSE,col.names = FALSE,sep = ",")

#unique genes:
unique_in_list0<-setdiff(genes_cluster0,c(genes_cluster1,genes_cluster2,genes_cluster3,genes_cluster4,genes_cluster))

```

```{r}
library(Seurat)
View(lung)
FeaturePlot(object = lung, features= c("CD68","GZMA","CXCR4"))

```


#8-CELL ANOTATION-------------------------------------------------------------------------------------------------

#1º convert the genes from the ensembl nomenclature to the gene symbol nomenclature:
```{r}
lung.norm@assays$RNA_Seq@counts

library(biomaRt)

# Connect to the Ensembl database
ensembl <- useMart("ensembl")

# Select the appropriate dataset (e.g., human genes)
ensembl <- useDataset("hsapiens_gene_ensembl",mart = ensembl)

# Convert Ensembl Gene IDs to gene names
ensembl_ids <- x@rowRanges@partitioning@NAMES  # Example Ensembl Gene IDs
gene_ensemble<-rownames(as.SingleCellExperiment(lung.norm))
gene_names <- getBM(attributes = c("ensembl_gene_id", "external_gene_name"),
                    filters = "ensembl_gene_id",
                    values = gene_ensemble,
                    mart = ensembl)

# Print the gene names
print(gene_names)

#########################
#EMPTY , ATTRIBUTE AS NA, AND IGNORE/REMOVE FROM THAT TABLE THE NA'S
gene_names[gene_names==""]<-NA
gene_names <- na.omit(gene_names)
View(gene_names)

your_single_cell_data_with_names = merge(as.SingleCellExperiment(lung.norm),gene_names,
                                         by.x = "ensembl_gene_id",
                                         by.y = "ensembl_gene_id",
                                         all.x = TRUE)

merged_df <- merge(as.SingleCellExperiment(lung.norm), gene_names)

rownames(lung.norm)
View(as.SingleCellExperiment(lung.norm))
  
merged_df <- merge(as.SingleCellExperiment(lung.norm),gene_names,by.x = "rowname", by.y = "ensembl_gene_id", all.x = TRUE)


#NEXT STEP SELECT THE OBJECT WITH THOSE ENSEMBLE GENES AND SUBSTITUTE WITH THE GENE NAMES.
```


#OPTION-A:Method 2-AUTOMATIC ASSIGMENT ANOTATIONS
```{r}
library(celldex)
ref <- BlueprintEncodeData()
ref
View(ref)

library(SingleR)

# predict cell types based on the reference
pred <- SingleR(test=as.SingleCellExperiment(lung), ref=ref, labels=ref$label.main)
table(pred$labels)
lung$singleR_annot = pred$labels
lung@assays$RNA@counts@Dimnames[[2]]=lung$singleR_annot
View(lung)

signature_matrix<-lung@assays$RNA@counts
write.csv(signature_matrix, "signature_matrix.csv")

marker_genes <-FindMarkers(lung, ident.1 = 2, min.pct = 0.25)

#COMPARE THE 2 TYPES OF CELL ANNOTATION:
#A-MANUAL ASSIGMENT
MANUAL_CELL_ANNOTATION<-DimPlot(lung, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
#B-REFERENCE BASED ASSIGNMENT
REFERENCE_CELL_ANNOTATION<-DimPlot(lung, reduction = "umap", label = TRUE, group.by="singleR_annot",pt.size = 0.5) + NoLegend()
MANUAL_CELL_ANNOTATION+REFERENCE_CELL_ANNOTATION

```

######################################################################################################
#OPTION-B:Method 1-Manual assignment: Using Marker Genes
```{r}
cluster1.markers <- FindMarkers(lung, ident.1 = 2, min.pct = 0.25)
head(cluster1.markers, n = 100)
```
#Compare one again the differentially expressed genes for cluster 5, comparativamente ao cluster 0 e 3.
```{r}
cluster5.markers <- FindMarkers(lung, ident.1 = 5, ident.2 = c(0, 3), min.pct = 0.25)
head(cluster5.markers, n = 5)
```

