---
title: "sc_lung_3 lung carcinomas"
output:
  pdf_document: default
  html_document: default
date: "2023-05-23"
---

#BIG NOTE:
#ALL THE GENES ARE IN ENSEMBL NOMENCLATURE, EXCEPT THE MITOCONDRIAL GENES
#THOSE ARE IN THE GENE SYMBOLS NOMENCLATURE

#libraries
```{r}
library(Seurat)
library(SeuratData)
library(gridExtra)
library(tidyverse)
library(data.table)
library(SingleCellExperiment)
library(scds)
library(useful)
library(ggplot2)

install.packages("Seurat")
install.packages("org.Hs.eg.db")
library(Seurat)
library(org.Hs.eg.db)
```

#DETERMINE THE DIRECTORY:
```{r}
knitr::opts_knit$set(root.dir = "C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/")

```


#load the data:
https://www.ebi.ac.uk/gxa/sc/experiments/E-MTAB-6653/downloads 
#1º:Load the cells
```{r}
library(readr)
mtx_cols <- read.table("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/barcodes.mtx_cols", header = FALSE, sep = "\t")
View(mtx_cols)

barcodes<-mtx_cols$V1
barcodes<-as.data.frame(barcodes,header=FALSE,rownames = FALSE)
file_path_<- file.path("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/", "barcodes.tsv.gz")
write.table(barcodes,file_path_,sep="\t",quote = FALSE,row.names = FALSE,col.names = FALSE)
View(barcodes)
```

#2º:Load the genes, convert into tsv and compress into .gz
```{r}
mtx_rows <- read.table("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/features.mtx_rows", header = FALSE, sep = "\t")
mtx_rows<-as.data.frame(mtx_rows$V1)
View(mtx_rows)

genes<-mtx_rows$`mtx_rows$V1`
genes<-as.data.frame(genes,header=FALSE,rownames = FALSE)
file_path <- file.path("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/", "features.tsv.gz")
write.table(genes,file_path,sep="\t",quote = FALSE,row.names = FALSE,col.names = FALSE)
View(genes)
```

#3º:TRY TO CONVERT THE ENSEMBL GENES INTO GENE SYMBOL AND DETERMINE THE MITOCONDRIAL GENES:
```{r}
#installations missing:
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("biomaRt")

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("org.Hs.eg.db")
library(ggplot2)
library(Matrix)
library(Seurat)
library(patchwork)
library(dplyr)
library(biomaRt)
library(org.Hs.eg.db)

#input list of the genes in the genes.tsv:
data_genes<-genes$genes

#BiomaRt method:
listEnsembl()
ensembl_<-useEnsembl(biomart = "genes")
datasets<-listDatasets(ensembl_)
ensembl <- useMart("ensembl", dataset="hsapiens_gene_ensembl")
attr<-listAttributes(ensembl)#here we define the reference to which type of conversion type we want to convert the ensembl genes
bm <- getBM(attributes=c("ensembl_gene_id", "external_gene_name"), values=data_genes, mart=ensembl)

# Check mitochondrial genes before conversion
print("Mitochondrial genes before conversion:")
print(grep("^MT-", data_genes, value = TRUE))#NO MITOCONDRIAL GENES BEFORE CONVERSION

#DETERMINE THE GENE SYMBOLS:
gene.symbols <- bm$external_gene_name[match(data_genes, bm$ensembl_gene_id)]
features_<-as.matrix(cbind(genes$genes,gene.symbols))
colnames(features_)<-c("ensembl_genes","gene_symbols")

#CREATE A NEW DATAFRAME WITH ENSEMBL GENES AND THE MITOCONDRIAL GENES IN GENE.SYMBOLS:
features<-as.data.frame(features_)
MT_GENES<-print(grep("^MT-",features_, value = TRUE))#13 MITOCONDRIAL GENES AFTER CONVERSION
MT_GENES<-as.list(MT_GENES)
features<- features[match(features$gene_symbols, MT_GENES, nomatch = 0) != 0,]

#CREATE THE FINAL FEATURES DATAFRAME:
features_df<-as.data.frame(features_)
#DO ONE BY ONE:
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198695"]<-"MT-ND6"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198712"]<-"MT-CO2"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198727"]<-"MT-CYB"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198763"]<-"MT-ND2"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198786"]<-"MT-ND5"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198804"]<-"MT-CO1"#
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198840"]<-"MT-ND3"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198886"]<-"MT-ND4"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198888"]<-"MT-ND1"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198899"]<-"MT-ATP6"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000198938"]<-"MT-CO3"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000212907"]<-"MT-ND4L"
features_df$ensembl_genes[features_df$ensembl_genes=="ENSG00000228253"]<-"MT-ATP8"


View(features_df)
#CREATE THE TSV FILE WITH GENES CALLED FEATURES.TSV
file_path <- file.path("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/", "features.tsv.gz")
write.table(features_df,file_path,sep="\t",quote = FALSE,row.names = FALSE,col.names = FALSE)
#1823 genes-not converted.gz
#22156 genes from the 23979 were actually converted

mtx_matrix<- read.table("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/matrix.mtx", header = FALSE, sep = "\t")
mtx_rows<-as.data.frame(mtx_rows$V1)
View(mtx_matrix)
file_path <- file.path("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/", "matrix.mtx.gz")
write.table(mtx_matrix,file_path,sep="\t",quote = FALSE,row.names = FALSE,col.names = FALSE)

```


#4º:ALL THE DATASET IN THE ACCORDING FORMATS ACCEPTED BY THE FUNCTION READ10X
```{r}
#gene.column = 1
lung.data <- Read10X(data.dir ="C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/E-MTAB-6653/single_cell_data_E_MTAB-6653/")
                     #,gene.column = 2,unique.features=TRUE)#try to change the genes here in the original data 23,979

feature_names <- rownames(lung.data)

# Identify features to keep (those without a dot followed by numbers)
features_to_keep <- !grepl("\\.\\d+", feature_names)

# Subset the Seurat object
lung_filtered <- lung.data[features_to_keep,]

#in the case, of gene names
lung_filtered <- lung.data[rownames(lung.data) != "",]

View(lung_filtered)
View(as.data.frame(lung_filtered))

#CREATE THE SEURAT OBJECT WITH THE GENE NAMES AND MINIMUM OF 3 CELLS AND 200 GENES:
lung<- CreateSeuratObject(counts = lung_filtered, project = "lung", min.cells = 3, min.features = 200)

dim(lung_data)
lung<-lung_data

#save para o diogo:
saveRDS(lung_data,file="C:/Users/ritas/Desktop/single_cell_Rita/lung_data.RDS")

View(lung)
head(rownames(lung),20)
View(lung@assays$RNA@counts@Dimnames)

############################################################################
#to join the cells automatic annotation to the data:
signature_matrix<-lung@assays[["RNA"]]@counts
write.table(signature_matrix, file="signature_matrix.txt", row.names=TRUE, col.names=TRUE,sep = "\t",quote = FALSE)
df = read.table("C:/Users/ritas/Desktop/TESE_IMM/SC_LUNG/SC_LUNG/signature_matrix.txt")
#23898 genes with gene names
############################################################################

```


#DETERMINE THE PERCENTAGE OF MITOCONDRIAL GENES ACCORDING TO THE PATTERN:
```{r}
#in the case of mouse genes
lung[["percent.mt"]] <- PercentageFeatureSet(lung, pattern = "^mt-") 
#OR
#in the case of human genes
lung[["percent.mt"]] <- PercentageFeatureSet(lung, pattern = "^MT-") 
lung[["percent.rp"]] <- PercentageFeatureSet(lung, pattern = "^RP(L|S)") 


VlnPlot(lung,features = c("nFeature_RNA","nCount_RNA","percent.mt","percent.rp"),ncol=4)

```

#OVERVIEW OF THE NFEATURE_RNA AND NCOUNT_RNA AND PERCENT.MT:
```{r}
View(lung@meta.data)
FeatureScatter(lung,feature1 = "nCount_RNA",feature2 = "nFeature_RNA")+geom_smooth(method = "lm")

#COMPARING PLOTS:
plot1 <- FeatureScatter(lung, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(lung, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
plot2
```

#FILTERING:
```{r}
lung<-subset(lung,subset=nFeature_RNA>200 & nFeature_RNA<5000 & percent.mt<25 & nCount_RNA<40000 & percent.rp<65)
#FIND DUBLETS:    [SOME AREAS THAT WERE MORE SEQUENCED, IN DUPLICATES]
lung_sce <- as.SingleCellExperiment(lung)
lung_sce <- cxds_bcds_hybrid(lung_sce, estNdbl=TRUE)#Function to find doublets/multiples in UMI
lung<- AddMetaData(
  object = lung,
  metadata = lung_sce$hybrid_call,
  col.name = "doublet"
)

table(lung_sce$hybrid_call)
lung<- subset(lung, subset = doublet == FALSE)

View(lung_sce)

#FALSE:30148 
#TRUE:1163   DOUBLETS
```

#3.Normalize data
```{r}
#1º:SEE THE MEANS AND VARIANCE OF THE DATA
lung<-lung_data
means <- sparseMatrixStats::rowMeans2(lung@assays$RNA@layers$counts)
variance <- sparseMatrixStats::rowVars(lung@assays$RNA@layers$counts)
plot(means, variance)
```

#2-A.TRY TO NORMALIZE THE DATA WITH LOG NORMALIZE:
```{r}
lung.norm<- NormalizeData(lung, normalization.method = "LogNormalize",scale.factor = 10000)

means <- sparseMatrixStats::rowMeans2(as.matrix(lung.norm@assays$RNA@layers$data))
variance <- sparseMatrixStats::rowVars(lung.norm@assays$RNA@layers$data)
plot(means, variance)

```

#2-B.TRY TO NORMALIZE THE DATA WITH SCTransform:
#NOTE:THE SCTRANSFORM TOOK A LONG TIME AND LEAD THE PC DOWN, SO I OPTED TO GO WITH THE CLASSIC NORMALIZATION:
```{r}
lung_sctranform <- SCTransform(lung,verbose=FALSE)

means_sctranform<- sparseMatrixStats::rowMeans2(lung_sctranform@assays$SCT@data)
variance_sctranform<- sparseMatrixStats::rowVars(lung_sctranform@assays$SCT@data)
plot(means_sctranform, variance_sctranform)

```


#3.SCALING (optional)------------------------------
```{r}
hist(sparseMatrixStats::colSums2(lung.norm@assays$RNA@layers$data), main = "Counts per Cell")
```

#4.IDENTIFY HIGHLY VARIABLE FEATURES (GENES)---------------------------------------
```{r}
nsclc <- FindVariableFeatures(lung.norm)

top10 <- head(VariableFeatures(nsclc), 10)

plot1 <- VariableFeaturePlot(nsclc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
plot2
plot1

```
#Identify the 10 most highly variable genes:
```{r}
top10<-head(VariableFeatures(nsclc),10)
top100<-head(VariableFeatures(nsclc),100)
print(top10)
print(top100)
```

#Plot variable features with and without labels:
```{r}
plot1<-VariableFeaturePlot(nsclc)
LabelPoints(plot=plot1,points = top10,repel = TRUE)#REPEL=TRUE, SHOWS THET POINTS WELL DISTINGUISH BETWEEN EACH OTHER.
```
##5.SCALING
REMOVE THE DESVIATIONS THAT LEAD TO THE CLUSTER OF THE CELLS, CAUSED BY THE DIFFERENCES ON CELL CYCLE OF THIS CELLS (BIOLOGICAL PROBLEM) OR THE BATCH EFFECT (TECNIQUAL PROBLEM); AND NOT BY THE BIOLOGICAL EFFECT.

```{r}
all.genes<-rownames(nsclc)
nsclc.scaled<-ScaleData(nsclc,features = all.genes)
str(nsclc.scaled)
#notes:
#normalized counts is stored in the data slot
#scaling data stored in scale.data
```
#B.DOWNSTREAM ANALYSIS
#6.Perform Linear dimensionality reduction------------------------------------------------------------------------------
```{r}
#6-A:TO PCA dimensionality reduction
ElbowPlot(lung)
lung<-RunPCA(nsclc.scaled,features = VariableFeatures(object = nsclc.scaled))
print(lung[["pca"]],dims=1:5,nfeatures=5)

VizDimLoadings(lung, dims = 1:2, reduction = "pca")

DimPlot(lung, reduction = "pca") + NoLegend()

FeaturePlot(lung,reduction = "pca",dims = c(1,2),features="percent.rp")

DimHeatmap(lung,dims=1,cells=500,balanced = TRUE)
DimHeatmap(lung,dims=1:10,cells=500,balanced = TRUE,nfeatures=50)
DimHeatmap(lung,dims=1:15,cells=500,balanced = TRUE)

```

```{r}
#6-B:FindNeighbors
#PARAMETERS:
#k.param: defines k for the k-nearest neighbor algorithm (depends- not a right number of k)
#reduction: Reduction to use as input for building the (S)NN
ElbowPlot(lung)

lung <- FindNeighbors(lung, dims = 1:20)

```

```{r}
#6-C:FindClusters
#PARAMETERS:
#RESOLUTION: Value of the resolution parameter, use a value above (below) 1.0 if you want to obtain a larger (smaller) number of communities.
#ex: resolution=0.5- 24 communities; resolution=0.8- 7communities
lung <- FindClusters(lung,resolution = seq(0.2, 0.7, 0.1))
#21 clusters

dim(lung)

# Look at cluster IDs of the first 5 cells
head(Idents(lung), 5)#obtain 34 levels
```

```{r}
#6-D:Plot the clusters:
p <- ggplot(data=as.data.table(lung@meta.data)[,.N,by="seurat_clusters"], mapping = aes(x=seurat_clusters, y=N, fill=seurat_clusters)) +
  geom_bar(stat="identity") +
  labs(x="Cluster Number", y="Number of Cells in that Cluster") +
  theme_classic()
print(p)
```

#UMAP VISUALIZATION:
```{r}
lung <- RunUMAP(lung, dims = 1:20)


#PLOT DIM:
#pt.size: Adjust point size for plotting
#shuffle: Whether to randomly shuffle the order of points. This can be useful for crowded plots if points of interest are being buried. (default is FALSE)
DimPlot(lung, reduction = "umap", label = TRUE,group.by = c("RNA_snn_res.0.2","RNA_snn_res.0.3","RNA_snn_res.0.4","RNA_snn_res.0.5","RNA_snn_res.0.6","RNA_snn_res.0.7"), label.box = FALSE)  & NoLegend() & NoAxes()

DimPlot(lung, reduction = "umap")

#The best resolution to the data is 0.6, attribute that to the data
Idents(object = lung) <-lung@meta.data$RNA_snn_res.0.6

FeaturePlot(lung, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rp"), order = TRUE) & NoLegend() & NoAxes()
```

#AUTOMATIC ANNOTATION- predict cell types based on the reference BlueprintEncodeData():
```{r}
#reference BlueprintEncodeData():
library(celldex)
ref <- BlueprintEncodeData()
ref

library(SingleR)
lung_counts<-GetAssayData(lung, slot="counts")

# predict cell types based on the reference
pred <- SingleR(test=lung_counts, ref=ref, labels=ref$label.main)
lung$singleR.labels<-pred$labels[match(rownames(lung@meta.data),rownames(pred))]

#Clustering with annotated cell type labels
DimPlot(lung,reduction = "umap",group.by = "singleR.labels")

table(pred$labels)
lung$singleR_annot = pred$labels
lung@assays$RNA@counts@Dimnames[[2]]=lung$singleR_annot
View(lung)

#Annotation diagnosis:
pred
pred$scores

plotScoreHeatmap(pred)
plotDeltaDistribution(pred)
